SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-11-05 2017-Xudong-SMT/ActaInformatica.tex
running ex, and decidability

\paragraph*{Term algebra.} \
Our models rely on a notion of parameterised actions that are
symbolic expressions using data types and variables. As we want to encode
the low-level behaviour of possibly very different
programming languages, we do not impose one specific algebra
for denoting actions, nor any specific communication mechanism. So we
leave unspecified the constructors of the algebra that will allow building
expressions and actions. Moreover, we use a generic {\em action interaction}
mechanism, based on unification between two or more action
expressions. This will be used in the semantics of synchronisation
vectors to express various kinds of communication or synchronisation mechanisms.

@@

In Table~\ref{table:BIPalgebra} we give an algebra presentation.
Sorts \texttt{Bool} and \texttt{Int} are predefined with standard operators.
Sort \texttt{Action} is similar, with a constructor \texttt{Synchro} denoting
a synchronised action, i.e. an ``internal'' action that cannot be
further synchronised with the environment. It also comes with an
overloaded \texttt{FUN} constructor, used to build actions with
arguments, which will be instantiated to the required sorts for a given
pNet.

@@

\QIN{
Let us first introduce a notation in LOTOS to give an intuition how we construct the algebra presentation.
}

\begin{example}[The ``Enable'' Operator]\label{example:EnableOperator}
\QIN{
Suppose there are two processes $P$, $Q$. In the Enable expression ``$P>>Q$'', an \texttt{exit} statement within $P$ terminates the current process, carrying a value $x$ that is captured by the \texttt{accept} statement of $Q$. 
}

\QIN{
A possible encoding of such ``Enable'' operator, called $EnableState$, uses a controller $CState$ with three simple control actions $l, r, \delta$ in it to synchronise the behaviour of $P$ and $Q$. Here we use a simple action algebra, containing two constructors $\delta(x)$ and $acc(x)$, for any possible data type of the variable $x$, corresponding to the statements \texttt{exit} and \texttt{accept}. To construct the algebra presentation, we extend the triple $\APres=\langle\mathit{Sorts},\mathit{Constrs},\mathit{Ops}\rangle$ shown in Table \ref{table:BIPalgebra} with four new constructors.}

\QIN{Using the \texttt{FUN} constructor, we express the custom constructors such as $\delta(x)$ in the form of $\texttt{FUN}_{Action, Data}(\delta, x)$ where $x$ is a variable of type $Data$. Here we have $arity(\texttt{FUN}_{Action, Data})=2$ and $\texttt{FUN}_{Action, Data}: Action, Data\to Action$. Note that the control action $\delta$ is present by the constructor $\delta$ of arity 0 here which is a different application of the constructor.
}

\QIN{
Another possible encoding of such ``Enable'' operator, called $EnableData$, uses a data-oriented style, with a single state controller $CData$, and a state-variable $v$, with values in $\{0,1\}$. As there is no more constructor required, we can use the same algebra presentation as $EnableState$ in this case.
}
\end{example}

Next, we enrich our example by composing two occurrences of the $EnableState$. In Section \ref{section:use-case:Enable} we will show how these encodings can be used to prove properties of the Enable operator.
\begin{example}[Composition of Two ``Enable'' Operators]\label{example:ComposeEnableOperators}
\QIN{
Suppose there are three processes $P$, $Q$ and $R$. In the expression ``$P>>(Q>>R)$'', an \texttt{exit} statement within $P$ terminates the current process, carrying a value $x_1$ that is captured by the \texttt{accept} statement of $Q$ in the expression $Q>>R$. And then, similarly, an \texttt{exit} statement within $Q$ terminates the current process, carrying a value $x_2$ that is captured by the \texttt{accept} statement of $R$.
}

\QIN{
We can compose two occurrences of the $EnableState$ to encode such expression. To distinguish the action of two controllers $CState1$ and $CState2$ used in such encoding $EnableStateCompLeft$, we names these control actions $l1, r1, \delta_1$ and $l2, r2, \delta_2$, respectively. The algebra presentation of $EnableStateCompLeft$ can be constructed by expending $\mathit{Constrs}$ with $\{\delta, acc, l1, l2, r1, r2, \delta_1, \delta_2\}$.
}
\end{example}

@@

A PLTS is a labelled transition system with variables, which can be
manipulated, defined, or accessed inside states, actions, guards, and
assignments. 
%
%Without loss of generality and to simplify the formalisation, we suppose 
%here that variables are local to 
%state: each state has its set of variables disjoint from the others. Transmitting 
%variable values from one state to the other can be done by explicit assignment. 
%
Each state has its set of variables called \emph{state variables}, 
which can only be modified by the assignment in transitions targeting this state. 
A global state variable of a PLTS is a state variable defined in all states.
% $x$ satisfying $\forall s\!\in\! S .\, x\!\in\! \vars(s)$. 
%
%Similarly, to simplify the management of variables and without loss of expressivity, we 
%suppose that transitions looping to the same state does not do assignments.
Note that we make no assumption on finiteness of the set of states, nor
on finite branching of the transition relation.

We first define the set of actions a PLTS can use.  Let $a$
range over action labels, $\symb{op}$ over operators, and $x_i$  over
variable names. Action terms are given by the following grammar:
\[
\begin{array}[l]{rcl@{\quad}p{5.5cm}}
\alpha\in\AlgA&::=&a(p_1,\ldots,p_n)&\text{action terms}\\
p_i&::=& ~\symb{Expr}&\text{parameters}\\
\symb{Expr}&::=& \symb{Value}~|~x~|~\symb{op}(\symb{Expr}_1,\dots,\symb{Expr}_n)&\text{expressions}
\end{array}
\]
%\[
%\begin{array}[l]{rcl@{\quad}p{5.5cm}}
%  \alpha\in\AlgA&::=&a(p_1,\ldots,p_n)&\text{action terms}\\
%  p_i&::=& ?x~|~\symb{Expr}&\text{parameters (input variable or expression)}\\
%  \symb{Expr}&::=& \symb{Value}~|~x~|~\symb{op}(\symb{Expr}_1,..,\symb{Expr}_n)&\text{Expressions}
%\end{array}
%\]

@@

\begin{example}[PLTS of the Controller in the ``Enable'' Operator]
\QIN{
The PLTS of the controller $CState$ of the encoding $EnableState$ defined in Example \ref{example:EnableOperator} is given as a triple $\langle S,s_0, \to\rangle$ where:
\begin{itemize}
	\item[$\bullet$] $S=\{0, 1\}$;
	\item[$\bullet$] $s_0=0$;
	\item[$\bullet$] $\to=\{0\xrightarrow{l}0, 0\xrightarrow{\delta}1, 1\xrightarrow{r}1\}$.
\end{itemize}
Add the guard and assignment to the label of the transition, the PLTS of the controller $CData$ is given as a triple $\langle S,s_0, \to\rangle$ where:
\begin{itemize}
	\item[$\bullet$] $S=\{0, 1\}$;
	\item[$\bullet$] $s_0=0$;
	\item[$\bullet$] $\to=\{0\xrightarrow{l, v=0}0, 0\xrightarrow{\delta,  v=0, \{v\leftarrow1\}}1, 1\xrightarrow{r, v=1}1\}$.
\end{itemize}
}
\end{example}
\QIN{The PLTSs contained in $EnableStateCompLeft$ are similar to the PLTS $CState$.}

@@

%% \begin{definition}[pNets]\label{def-pnets}
%% A pNet is a hierarchical structure, whereof the leaves are PLTSs and holes:
%% $\pNet::= PLTS~|~\langle \pNet_i^{i\in I}, S_j^{j\in J}, \symb{SV}_k^{k\in K}\rangle$,
%% where
%% \begin{itemize}
%% \item[$\bullet$] %\noteEM{Pire que ca, $\I$ n'est pas
%%                  %defini...}\noteSB{Why is one $\I$ and the other
%%   %$\I_\P$?}
%%   $I$ and $J$ are the sets over which are indexed sub-pNets and holes, respectively, such that: $I\cap J=\emptyset$ and  $I\cup J\neq\emptyset$.

%% \item[$\bullet$] $\pNet_i^{i\in I}$ is a family of sub-pNets.

%% \item[$\bullet$] $S_j \subseteq \AlgA$ is a set of action terms,
%%   denoting the $\Sort$\footnote{The definition of $Sorts$ (the
%%     set of actions of a Hole  
%%   or pNet), and also of $Leaves$ and $Holes$ (respectively all PLTSs
%%   and all holes in a pNet), can be found in \cite{henrio:Forte2016}.}
%%   of hole $j$.

%% \item[$\bullet$] $\symb{SV}_k^{k\in K}$, with $K\in\I_\P$, is a set of
%%   synchronisation vectors: for each $k\!\in\! K$,
%%   $\symb{SV}_k\!=\!\alpha_{l}^{l\in I_k \uplus J_k}\to\alpha'_k$ where
%%   $\alpha'_k\in \mathcal{A}_\P$, $I_k\subseteq I$, $J_k\subseteq J$,
%%   $\forall i\!\in\! I_k.\,\alpha_{i}\!\in\!\Sort(\pNet_i)$
%% %  \noteSB{What is the sort of a pNet? Should we define $\Sort(\pNet)
%% %    \triangleq \{\Label(\symb{SV}_k\,|\,k \in K\}$?}
%%   , 
%%   $\forall j\!\in\!
%%   J_k.\,\alpha_{j}\!\in\!S_j$, and $\vars(\alpha'_k)\subseteq \bigcup_{l\in I_k\uplus 
%%   J_k}{\vars({\alpha_l})}$. The global action of a vector $\symb{SV}_k$ is
%%   $\Label(\symb{SV}_k) = \alpha'_k$.
%% \end{itemize}
%% \end{definition}

@@

\begin{example}[pNet of the ``Enable'' Operator]\label{example:pNetEnable}
\QIN{
Consider the encoding $EnableState$ defined in Example \ref{example:EnableOperator}, the processes $P$ and $Q$ are holes of the pNet.} In this example we use \emph{synchronised actions}, to express actions that cannot be further synchronised, similar to the silent action $\tau$ of CCS. We denote them as any action expression with the text underlined, as e.g. $\underline{\delta(x_1)}$. Such synchronised actions do not play any special role for defining the semantics of pNets, but as one can expect, will be crucial for defining weak equivalences. \QIN{The pNet can be given as follow:
\begin{itemize}
	\item[$\bullet$] $\pNet=\{CState\}$;
	\item[$\bullet$] $J=\{P, Q\}$;
	\item[$\bullet$] $\symb{SV}=\{\langle a_1, \text{-}, l\rangle\rightarrow a_1[\forall y1.a_1\neq \delta(y1)], 
	\langle \delta(x_1), acc(x_1), \delta\rangle\rightarrow\underline{\delta(x_1)},\\
	\langle \text{-}, a_2, r\rangle\rightarrow a_2[\forall y2.a_2\neq acc(y2)]\}$.
\end{itemize}
Replace $CState$ with $CData$ the pNet of $EnableData$ can be given with similar $\symb{SV}$.
}
\end{example}

Note that synchronisation vectors are defined in a parameterised
manner:  the first and third lines represent one vector
for each possible action of hole $P$
(resp. $Q$). This notation can also use predicates, as in the first
case, in which we want the vector to apply to any action of $P$
except for $\delta(z)$.

\QIN{The pNet of $EnableStateCompLeft$ presents a bigger tree with three levels containing a pNet as its node.}
\begin{example}[pNet of the Composition of ``Enable'' Operators]\label{example:pNetComposeEnable}
\QIN{
Consider the encoding $EnableStateCompLeft$ defined in Example \ref{example:ComposeEnableOperators}, the processes $P$, $Q$ and $R$ are holes of the pNet. Let $pNet_1$ be the sub-pNet that presents $Q>>R$. The pNet can be given as follow:
\begin{itemize}
	\item[$\bullet$] $\pNet=\{pNet_1, CState1\}$;
	\item[$\bullet$] $J=\{P\}$;
	\item[$\bullet$] $\symb{SV}=\{\langle a_1, \text{-}, l1\rangle\rightarrow a_1[\forall y1.a_1\neq \delta(y1)], 
	\langle \delta(x_1), acc(x_1), \delta\rangle\rightarrow\underline{\delta(x_1)},\\
	\langle \text{-}, a_2, r1\rangle\rightarrow a_2[\forall y2.a_2\neq acc(y2)]\}.$
\end{itemize}
The sub-pNet $pNet_1$ is a pNet of a single ``Enable'' operator as the pNet in Example \ref{example:pNetEnable}. Actions in $pNet_1$ should be renamed to distinguish themselves from those already used in the pNet of $EnableStateCompLeft$.
}
\end{example}

In Figures~\ref{schema:enable-pnets},\ref{schema:enable-composed}, we show
examples of these constructs, with two PLTSs, one hole and one pNet
node encoding our running example.

@@

  \label{OT:example-2}
        \begin{mathpar}
	  OT_2'  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{\delta(x2)}_P ~~
          \xrightarrow{acc(x2)}_Q \} ~~
          [v_1=0 \land v_2 = 0 \land \forall z. acc(x2) \neq
            \delta(z)] ~~ \{v1:=1\}
        }
	{\ostate{0,0} \xrightarrow {\underline{\delta(x2)}} \ostate{0,0}}
	\end{mathpar}
        where $v_1$ and $v_2$ are respectively the state variables of
        the first and second $EnableData$ pNets. Indeed the global
        state space of this version is reduced to $\ostate{0,0}$, and
        the state change is replaced by the assignment ``$v_1:=1$''.

@@

\begin{example}[Unsat OT]
  \label{schema:unsat-ot}
\begin{mathpar}
	  OT_{19}  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{a1}_P ~~
          \xrightarrow{acc(x2)}_R \} ~~
          [a1 = \delta(x2) \land \forall z. a1 \neq \delta(z)]
        }
	{\ostate{2,0} \xrightarrow {\underline{\delta(x2)}} \ostate{3,0}}
	\end{mathpar}

Here we display an  
unsatisfiable open transition 
from the open automaton of our running example.
It shows the case where processes $P$ and $R$ ``try'' to communicate directly, but this is made impossible by the combination of chosen synchronisation vectors and their guards. This
  mismatch is materialised by the predicate
fragment ``$a1 = \delta(x2) \land \forall z. a1 \neq \delta(z)$''. In Section~\ref{section:TranslationToSMTlib} we will show how this is encoded and checked using the SMT engine.
\end{example}

@@

In Algorithm~\ref{alg1}, functions \textit{filterUnsatTransitions} (in line 11) 
\textit{RefineReachableSubAutomaton} (in line 13) make use of the Z3 engine to implement the
satisfiability checking, for all the OTs computed that far.
For a transition
\[		\inferrule*[myfraction=\reddottedrule]
		{%\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
			\{\xrightarrow{\beta_j}_j\}^{j\in J}, \Pred\,, \Post}
		{s \xrightarrow {\alpha}s'}
                \]
                we need to build a predicate encoding both the OT predicate and
the information available on the possible assignments of the initial
state $s$. This is expressed as
\[Pred_{l}\wedge \bigwedge_{v\in vars(s)}\{ \bigvee_{Assign\in
  assigns(v,s)} Assign\}
\]
where each $Assign$ represents one of the possible (symbolic) values of
 the state variable $v$.

@@

\subsection{Translation to SMTlib}
\label{section:TranslationToSMTlib}

@@

Figure~\ref{schema:smt-lib} shows the translation of  Figure~\ref{schema:unsat-ot}
in the SMTlib syntax. It contains the declaration
of the relevant part of the LOTOS action algebra sorts 
and constructors, then the declaration of variables, and finally the
predicate to be checked, encoded as a set of assertions.

Here the behavior of the hole ``$P$'' ($a1$ in Figure~\ref{schema:unsat-ot}), after matching with the synchronisation vectors, appears as a variable
``|a1:sva\_SVA0:15:1|'', and ``$R$'' ($acc(x2)$) as ``|(ACT delta
|x:sva\_SV4:1:2|''. Line 16 requires that $a1 = \delta(x2)$ to match
the second synchronisation vector of the root pNet node, and lines
17-18 that $a1$ does not match $\delta(x1)$ for any $x1$. 
We also display  the diagnosis (``SAT'' or ``UNSAT'') generated by Z3.

>>>>

\paragraph*{Term algebra.} \
Our models rely on a notion of parameterised actions that are
symbolic expressions using data types and variables. As we want to encode
the low-level behaviour of possibly very different
programming languages, we do not impose one specific algebra
for denoting actions, nor any specific communication mechanism. So we
leave unspecified the constructors of the algebraused to build
expressions and actions. Moreover, we use a generic {\em action interaction}
mechanism, based on unification between two or more action
expressions. This will be used in the semantics of synchronisation
vectors to express various kinds of communication or synchronisation mechanisms.

@@

In Table~\ref{table:BIPalgebra} we give an algebra presentation.
Sorts \texttt{Bool} and \texttt{Int} are predefined with standard operators.
Sort \texttt{Action} is similar, with a constructor \texttt{Synchro} denoting
a synchronised action, i.e. an ``internal'' action that cannot be
further synchronised with the environment. It also comes with an
overloaded \texttt{FUN} constructor, used to build action expressions with
arguments, which will be instantiated to the required sorts for a given
pNet.

@@

\subsection{Running example: the Enable operator}
\ERIC{To illustrate our definitions and results, we will use a small running example, coming from the specification language LOTOS \cite{LotosISO89}.
  Suppose we have two unspecified processes $P$ and $Q$. In the Enable expression ``$P>>Q$'', an \texttt{exit} statement within $P$ terminates the current process and pass the control to $Q$, carrying a value $x$ that is captured by the \texttt{accept} statement of $Q$.
  In the next section we will define pNet encodings of the Enable operator. Let us start by defining the algebra presentation that will be used in these pNets.
}

\begin{example}[The ``Enable'' Operator]\label{example:EnableOperator}
\ERIC{In Fig. \ref{schema:enable-pnets} we shall give two possible pNet  encodings of the ``Enable'' operator respectively state-oriented and data-oriented. 
The first one is called $EnableState$, and uses a controller $CState$ with three simple control actions $l, r, \delta$ in it to synchronise the behaviour of $P$ and $Q$. Here we use a simple action algebra, containing two constructors $\delta(x)$ and $acc(x)$, for any possible data type of the variable $x$, corresponding to the LOTOS statements \texttt{exit} and \texttt{accept}. To construct the algebra presentation, we extend the triple $\APres=\langle\mathit{Sorts},\mathit{Constrs},\mathit{Ops}\rangle$ shown in Table \ref{table:BIPalgebra} with four new action constant constructors $l, r, \delta, acc$.}

\ERIC{This shows an example of the generic ``FUN'' constructor used to construct expressions (of the Action sort, or of any data sort) in any algebra presentations.
  Here \texttt{FUN} will be instanciated e.g. as, $\delta(x)$ in the form of $\texttt{FUN}_{Action, Data}(\delta, x)$ where $x$ is a variable of type $Data$.
  %Here we have $arity(\texttt{FUN}_{Action, Data})=2$ and $\texttt{FUN}_{Action, Data}: Action, Data\to Action$.
  Note that the same action constructor $\delta$ is also used as a constant action in the controller $CState$.
}

\ERIC{The $EnableData$ pNet will use the same algebra presentation.}
%\QIN{
%Another possible encoding of such ``Enable'' operator, called $EnableData$, uses a data-oriented style, with a single state controller $CData$, and a state-variable $v$, with values in $\{0,1\}$. As there is no more constructor required, we can use the same algebra presentation as $EnableState$ in this case.
%}
\end{example}

@@

A PLTS is a labelled transition system with variables, which can be
manipulated, defined, or accessed inside states, actions, guards, and
assignments. 
%
%Without loss of generality and to simplify the formalisation, we suppose 
%here that variables are local to 
%state: each state has its set of variables disjoint from the others. Transmitting 
%variable values from one state to the other can be done by explicit assignment. 
%
Each state has its set of variables called \emph{state variables}, 
which can only be modified by the assignment in transitions targeting this state. Variables of the initial state must be initialized.
%Eric: I have removed the mention of ``global variables''. Do we use them somewhere else in the paper ? I did not find any.}
%A global state variable of a PLTS is a state variable defined in all states.
% $x$ satisfying $\forall s\!\in\! S .\, x\!\in\! \vars(s)$. 
%
%Similarly, to simplify the management of variables and without loss of expressivity, we 
%suppose that transitions looping to the same state does not do assignments.
Note that we make no assumption on finiteness of the set of states, nor
on finite branching of the transition relation.

We first define the set of actions a PLTS can use.  Let $a$
range over action constants, $\symb{op}$ over operators, and $x$  over
variable names. Action terms are given by the following grammar:
\[
\begin{array}[l]{rcl@{\quad}p{5.5cm}}
\alpha\in\AlgA&::=&a(p_1,\ldots,p_n)&\text{action terms}\\
p_i&::=& ~\symb{Expr}&\text{parameters}\\
\symb{Expr}&::=& \symb{Value}~|~x~|~\symb{op}(\symb{Expr}_1,\dots,\symb{Expr}_n)&\text{expressions}
\end{array}
\]
%\[
%\begin{array}[l]{rcl@{\quad}p{5.5cm}}
%  \alpha\in\AlgA&::=&a(p_1,\ldots,p_n)&\text{action terms}\\
%  p_i&::=& ?x~|~\symb{Expr}&\text{parameters (input variable or expression)}\\
%  \symb{Expr}&::=& \symb{Value}~|~x~|~\symb{op}(\symb{Expr}_1,..,\symb{Expr}_n)&\text{Expressions}
%\end{array}
%\]

@@

\begin{example}[PLTS of the Controller in the ``Enable'' Operator]
\ERIC{
The PLTS of the controller $CState$ from Example \ref{example:EnableOperator} is  $\langle S,s_0, \to\rangle$, where:
\begin{itemize}
	\item[$\bullet$] $S=\{0, 1\}$, $s_0=0$;
	\item[$\bullet$] $\to=\{0\xrightarrow{l}0, 0\xrightarrow{\delta}1, 1\xrightarrow{r}1\}$.
\end{itemize}
Featuring guards and assignments in the label of transitions, the PLTS of the controller $CData$ is $\langle S,s_0, \to\rangle$ where:
\begin{itemize}
	\item[$\bullet$] $S=\{0, 1\}$, $s_0=0$, $vars(s_0)=\{v\}$, $\texttt{init}\ \{v:=0\}$;
	\item[$\bullet$] $\to=\{0\xrightarrow{l, [v=0]}0, 0\xrightarrow{\delta,  [v=0], \{v\leftarrow1\}}1, 1\xrightarrow{r, [v=1]}1\}$.
\end{itemize}
}
\ERIC{The PLTSs contained in $EnableStateCompLeft$ are similar to the PLTS $CState$, but their action constants have been renamed, ensuring disjoint state variable sets.}
\end{example}

@@

@@

%% Next, we enrich our example by composing two occurrences of the $EnableState$. In Section \ref{section:use-case:Enable} we will show how these encodings can be used to prove properties of the Enable operator.
%% \begin{example}[Composition of Two ``Enable'' Operators]\label{example:ComposeEnableOperators}
%% \QIN{
%% Suppose there are three processes $P$, $Q$ and $R$. In the expression ``$P>>(Q>>R)$'', an \texttt{exit} statement within $P$ terminates the current process, carrying a value $x_1$ that is captured by the \texttt{accept} statement of $Q$ in the expression $Q>>R$. And then, similarly, an \texttt{exit} statement within $Q$ terminates the current process, carrying a value $x_2$ that is captured by the \texttt{accept} statement of $R$.
%% }
%% \QIN{
%% We can compose two occurrences of the $EnableState$ to encode such expression. To distinguish the action of two controllers $CState1$ and $CState2$ used in such encoding $EnableStateCompLeft$, we names these control actions $l1, r1, \delta_1$ and $l2, r2, \delta_2$, respectively. The algebra presentation of $EnableStateCompLeft$ can be constructed by expending $\mathit{Constrs}$ with $\{\delta, acc, l1, l2, r1, r2, \delta_1, \delta_2\}$.
%% }
%% \end{example}

\begin{example}[pNet of the ``Enable'' Operator]\label{example:pNetEnable}
\ERIC{
  Consider the pNet $EnableState$ shown in Fig. \ref{schema:enable-pnets}, upper part. It has two holes $P$ and $Q$, and one PLTS $CState$. Note that our graphical syntax presents pNet nodes and synchronisation vectors as finite arrays, where the index sets follow the positions in the arrays.
  In this example we use \emph{synchronised actions}, to express actions that cannot be further synchronised, similar to the silent action $\tau$ of CCS. We denote them as any action expression but with the text underlined, as e.g. $\underline{\delta(x_1)}$. Such synchronised actions do not play any special role for defining the semantics of pNets, but as one can expect, will be crucial for defining weak equivalences. The pNet can be given as follow:
\begin{itemize}
	\item[$\bullet$] $\pNet=\{CState\}$;
	\item[$\bullet$] $J=\{P, Q\}$;
	\item[$\bullet$] $\symb{SV}=$ as listed on the right side of each pNet node.
        %%   \{\langle a_1, \text{-}, l\rangle\rightarrow a_1[\forall y1.a_1\neq \delta(y1)], 
	%% \langle \delta(x_1), acc(x_1), \delta\rangle\rightarrow\underline{\delta(x_1)},\\
	%% \langle \text{-}, a_2, r\rangle\rightarrow a_2[\forall y2.a_2\neq acc(y2)]\}$.
\end{itemize}
Consider the first  vector of $EnableState$, written as $\langle a_1, \text{-}, l\rangle\rightarrow a_1[\forall y1. a_1\neq \delta(y1)]$. It denotes a possible behavior of $P>>Q$, for any possible action of the hole $P$ that respects the vector predicate, that is any action that does not match $\delta(y1)$, for any value of the variable $y1$.
Remark that the data-oriented encoding, in pNet $EnableData$ has the same synchronisation vectors, upto alpha conversion. Only the controller $CState$ has been replaced by $CData$.
}
\end{example}

%% Note that synchronisation vectors are defined in a parameterised
%% manner:  the first and third lines represent one vector
%% for each possible action of hole $P$
%% (resp. $Q$). This notation can also use predicates, as in the first
%% case, in which we want the vector to apply to any action of $P$
%% except for $\delta(z)$.

%\QIN{The pNet of $EnableStateCompLeft$ presents a bigger tree with three levels containing a pNet as its node.}
\begin{example}[pNet of the Composition of ``Enable'' Operators]\label{example:pNetComposeEnable}
\ERIC{
  Now consider the pNet $EnableStateCompLeft$ in Fig. \ref{schema:enable-composed}; the processes $P$, $Q$ and $R$ are holes of the pNet.
It has been obtained by replacing the second hole in pNet $EnableState$ by another pNet representing $Q>>R$. Technicaly, the action constants in the PLTS $CState1$ and $CState2$ have been renamed to avoid clashes, and the variables in the vectors also given different names.
%%   Let $pNet_1$ be the sub-pNet that presents $Q>>R$. The pNet can be given as follow:
%% \begin{itemize}
%% 	\item[$\bullet$] $\pNet=\{pNet_1, CState1\}$;
%% 	\item[$\bullet$] $J=\{P\}$;
%% 	\item[$\bullet$] $\symb{SV}=\{\langle a_1, \text{-}, l1\rangle\rightarrow a_1[\forall y1.a_1\neq \delta(y1)], 
%% 	\langle \delta(x_1), acc(x_1), \delta\rangle\rightarrow\underline{\delta(x_1)},\\
%% 	\langle \text{-}, a_2, r1\rangle\rightarrow a_2[\forall y2.a_2\neq acc(y2)]\}.$
%% \end{itemize}
%% The sub-pNet $pNet_1$ is a pNet of a single ``Enable'' operator as the pNet in Example \ref{example:pNetEnable}. Actions in $pNet_1$ should be renamed to distinguish themselves from those already used in the pNet of $EnableStateCompLeft$.
}
\end{example}

%% In Figures~\ref{schema:enable-pnets},\ref{schema:enable-composed}, we show
%% examples of these constructs, with two PLTSs, one hole and one pNet
%% node encoding our running example.

@@

  \label{OT:example-2}
        \begin{mathpar}
	  OT_2'  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{\delta(x2)}_P ~~
          \xrightarrow{acc(x2)}_Q \} ~~
          [v_1=0 \land v_2 = 0 \land \forall y1. acc(x2) \neq
            \delta(y1)] ~~ \{v1:=1\}
        }
	{\ostate{0,0} \xrightarrow {\underline{\delta(x2)}} \ostate{0,0}}
	\end{mathpar}
        where $v_1$ and $v_2$ are respectively the state variables of
        the first and second $EnableData$ pNets. Indeed the global
        state space of this version is reduced to $\ostate{0,0}$, and
        the state change is replaced by the assignment ``$v_1:=1$''.

@@

\begin{example}[Unsat OT]
  \label{schema:unsat-ot}
\begin{mathpar}
	  OT_{19}  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{a1}_P ~~
          \xrightarrow{acc(x2)}_R \} ~~
          [a1 = \delta(x2) \land \forall y1. a1 \neq \delta(y1)]
        }
	{\ostate{2,0} \xrightarrow {\underline{\delta(x2)}} \ostate{3,0}}
	\end{mathpar}

Here we display an  
unsatisfiable open transition 
from the open automaton of Fig. \ref{schema:enable-composed}.
It shows the case where processes $P$ and $R$ ``try'' to communicate directly, but this is made impossible by the combination of chosen synchronisation vectors and their guards. This
  mismatch is materialised by the predicate
fragment ``$a1 = \delta(x2) \land \forall z. a1 \neq \delta(z)$''. In Section~\ref{section:TranslationToSMTlib} we will show how this is encoded and checked using the SMT engine.
\end{example}

@@

In Algorithm~\ref{alg1}, functions \textit{filterUnsatTransitions} (in line 11) 
\textit{RefineReachableSubAutomaton} (in line 13) make use of the Z3 engine to implement the
satisfiability checking, for all the OTs computed that far.
For a transition
\[		\inferrule*[myfraction=\reddottedrule]
		{%\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
			\{\xrightarrow{\beta_j}_j\}^{j\in J}, \Pred\,, \Post}
		{s \xrightarrow {\alpha}s'}
                \]
                we need to build a predicate encoding both the OT predicate and
the information available on the possible assignments of the initial
state $s$. This is expressed as
\[Pred_{l}\wedge \bigwedge_{v\in vars(s)}\{ \bigvee_{Assign\in
  assigns(v,s)} Assign\}
\]
where each $Assign$ represents one of the possible (symbolic) values of
variable $v$.

\ERIC{Each call to the SMT engine will return either SAT or UNSAT results, or may fail due to time bounds. Also, the accuracy of a 'SAT' result may depend on the characteristics of the specific theory used to axiomatise the data domains and operators. Such theories may be intrinsically undecidable, or may not properly handled by the SMT engine (e.g. for recursive datatypes).
However, while building the open automaton, that is a symbolic representation of the pNet behavior, it is not logically incorrect to keep potentialy unsatisfiable transitions, so we will treat undecidable results, and time bounds failures, as SAT. So the open automaton, may have more transitions and states that the theoretical minimum.}

\ERIC{So this is not a problem in itself when constructing the automaton, but a similar limitation will arise later when using open automata as the input of model checking or equivalence checking algorithms. Note that decidability results for these 
will be dependent on the decidability of the SMT theories used, as is proved e.g. in \cite{henrio:Forte2016} for bisimulation.}

@@

\subsection{Translation to SMTlib and Satisfiability Check}
\label{section:TranslationToSMTlib}

@@

Figure~\ref{schema:smt-lib} shows the translation of  Example~\ref{schema:unsat-ot}, page \pageref{schema:unsat-ot},
in the SMTlib syntax. It contains the declaration
of the relevant part of the LOTOS action algebra sorts 
and constructors, then the declaration of variables, and finally the
predicate to be checked, encoded as a set of assertions.

Here the behavior of the hole ``$P$'' ($a1$ in Example~\ref{schema:unsat-ot}), after matching with the synchronisation vectors, appears as a variable
``|a1:sva\_SVA0:15:1|'', and ``$R$'' ($acc(x2)$) as ``|(ACT delta
|x:sva\_SV4:1:2|''. Line 16 requires that $a1 = \delta(x2)$ to match
the second synchronisation vector of the root pNet node, and lines
17-18 that $a1$ does not match $\delta(x1)$ for any $x1$. 
We also display  the diagnosis (``SAT'' or ``UNSAT'') generated by Z3.
